Is Archaeology a science? Insights and imperatives from 10,000 articles and a year of reproducibility reviews

### 2025-05-08 analysis/paper/paper.qmd
discussing qualitative research

The question of archaeology's status as a science usually comes up in the context of what the discipline should or should not be. One of the first landmarks in tackling this question is the debate published by *Antiquity* between classical archaeologist Jacquetta Hawkes and palaeoanthropologist Glynn Isaac. @hawkesProperStudyMankind1968, advocating a humanistic archaeology, was concerned that scientific approaches to archaeology were causing researchers to be "swamped by a vast accumulation of insignificant, disparate facts, like a terrible tide of mud, quite beyond the capacity of any man to contain and mould into historical form". More optimistic about the integration of science and archaeology, @isaacWhitherArchaeology1971 counters that "New levels of precision in presenting data and in interpreting them can surely lead to briefer and more interesting technical reports as well as providing the basis for more lively literary portrayals of what happened in prehistory. Expanding on Isaac's perspective, @binford1962archaeology argued that archaeology should operate as a science after the model proposed by philosopher Carl Hempel, which prescribed hypothesis-driven approaches, leading to generalizable laws of human behavior. Drawing on a different group of philosophers, @Smith_2017 argues for archaeology more specifically as a social science. Counter-arguments to archaeology as a science come from numerous directions, notably @hodder1985postprocessual who rejected the quest for generalisations and instead argued that archaeology should be subjective and reflective, focussed on symbolic and relational meanings of material culture and the historical particularity of past human cultures. These debates, and the many more similar ones summarised by @martinon-torresArchaeologicalTheoriesArchaeological2013, have become a genre in archaeological writing that can be characterized as mostly based on personal observations, microscopic dissections of a handful of cherry-picked case studies of good or bad practice, and discussion of various philosophers and sociologists.

@@

To objectively quantify the diversity of modern archaeological practice across a scale of relative hardness or softness, as an evaluation of its status as a science, and the place of this journal in context of other archaeology journals, I take a bibliometric approach. This approach is based on @fanelliBibliometricEvidenceHierarchy2013, who examined the hardness and softness of 12 disciplines using scholarly publication parameters. @fanelliBibliometricEvidenceHierarchy2013 found a spectrum of statistically significant variation in bibliometric variables from the physical to the social sciences, with papers tending at the softer end of the spectrum tending to have fewer co-authors, use less substantive titles, have longer texts, cite older literature, and have a higher diversity of sources. In @fanelliBibliometricEvidenceHierarchy2013's analysis harder sciences include Space Science, Physics, Chemistry, less hard sciences include social sciences include Psychiatry, Psychology, Economics, Business and General Social Sciences, and the Humanities define the soft end of the spectrum. Following @fanelliBibliometricEvidenceHierarchy2013, I quantify the number of authors, length of article, relative title length, age of references, and diversity of references for a large sample of peer-reviewed journal articles.

@@

With many disciplines making increasing use of computationally intensive analyses in recent years there has been renewed interest in reproducibility [@levequeReproducibleResearchScientific2012]. In part, this is because computationally intensive research is difficult to communicate within the constraints of the methods section of a traditional journal article — the reader also needs the computer code written by the original authors, not just the article text. There is also the broader context of rising pressure to publish in prestigious journals and intense competition for funds that create strong incentives for malpractice in research [@edwards2017academic]. These two factors have led to widespread concerns of a reproducibility crisis in many fields [@Baker2016]. Estimates of scientific reproducibility in several fields confirm the extent of this problem. Empirical replications of 100 studies published in three psychology journals found that 36% of replications had statistically significant results, compared to 96% of the original studies [@OpenScienceCollaboration2015]. Similar empirical replications of large numbers of social science studies and experiemental economics studies successfully replicated 61% and 62% of their target studies respectively [@camererEvaluatingReplicabilitySocial2018; @camererEvaluatingReplicabilityLaboratory2016].

@@

How does archaeology compare to these other fields in terms of reproducibility? Empirical reproducibility has long been valued in field archaeology. Throughout the history of archaeology, well-known sites have been repeatedly revisited to test old hypotheses with new evidence or methods, for example, Olduvai Gorge (Tanzania), Cahokia (USA), Çatalhöyük (Turkey), and Madjedbebe (Australia). Similarly among many experimental archaeologists, empirical reproducibility is a key concern, for example in lithic use-wear identification [@hayesLearningBlindTests2017] and the measurement of lithics [@pargeterReplicabilityLithicAnalysis2023]. To what extent does this concern extend to computational reproducibility among archaeologists?

@@

library(patchwork)
(aer_plot_software + aer_plot_methods + aer_plot_repo) / (aer_plot_issues + aer_plot_issues_lang) + plot_annotation(tag_levels = 'A')

@@

Despite the relatively small number of reproducibility reviews reported on here, there are patterns of common issues that point to a small set of simple tasks authors can do that have high potential to increase reproducibility. The problem of incomplete materials can be tackled in several basic ways. First, authors should use a simple and logical folder structure to organise their code and data to be as self-contained as possible (ie. not requiring files outside of the project folder). There are many excellent, simple, and widely-used templates that authors can choose from, e.g. @doi:10.1080/00031305.2017.1375986, @figueiredoSimpleKitUse2022, @Cookiecutter, @cooperGuideReproducibleCode2017 and @wilsonGoodEnoughPractices2017

@@

@@

Computational reproducibility is not a panacea, results that are fully reproducible can contain errors and fraud. It is no guarantee of code quality, or that statistics have been used appropriately (cf. @Vaiglova_2025; Crema 2025), or that data management is consistent with FAIR and CARE principles [@carrollOperationalizingCAREFAIR2021]. It is also time-consuming for authors to ensure their computational work can be reproduced, and for reviewers to evaluate. In a professional environment where job security and career progression is often associated with pressure to publish many high-impact papers, demands for authors to spend time on reproducibility, resulting in less time for publishing more papers, may seem frustrating. This may seem especially unfair to early career researchers on short-term contracts, who may feel the goalposts for career success are being moved and that they are being asked to do more work their graduate training has not prepared them for. This highlights the need for a culture shift among senior archaeological scientists to value reproducibility in hiring and promotion decision-making. Professors must contribute to this shift by nurturing a culture of reanalysis and reproducibility in their teaching, for example by using replication assignments and by training students in the most current best practices and tools for reproducible research, such as R and Python [@marwickHowUseReplication2020].

>>>>

The question of archaeology's status as a science usually comes up in the context of what the discipline should or should not be. One of the first landmarks in tackling this question is the debate published by *Antiquity* between classical archaeologist Jacquetta Hawkes and palaeoanthropologist Glynn Isaac. @hawkesProperStudyMankind1968, advocating a humanistic archaeology, was concerned that scientific approaches to archaeology were causing researchers to be "swamped by a vast accumulation of insignificant, disparate facts, like a terrible tide of mud, quite beyond the capacity of any man to contain and mould into historical form". More optimistic about the integration of science and archaeology, @isaacWhitherArchaeology1971 counters that "New levels of precision in presenting data and in interpreting them can surely lead to briefer and more interesting technical reports as well as providing the basis for more lively literary portrayals of what happened in prehistory. Expanding on Isaac's perspective, @binford1962archaeology argued that archaeology should operate as a science after the model proposed by philosopher Carl Hempel, which prescribed hypothesis-driven approaches, leading to generalizable laws of human behavior. Drawing on a different group of philosophers, @Smith_2017 argues for archaeology more specifically as a social science. @bevan2015data proposes that floods of digital data are reconfiguring our analytical agendas and support empirical and inductive inference. Counter-arguments to archaeology as a science come from numerous directions, notably @hodder1985postprocessual who rejected the quest for generalisations and instead argued that archaeology should be subjective and reflective, focussed on symbolic and relational meanings of material culture and the historical particularity of past human cultures. These debates, and the many more similar ones summarised by @martinon-torresArchaeologicalTheoriesArchaeological2013, have become a genre in archaeological writing that can be characterized as mostly based on personal observations, microscopic dissections of a handful of cherry-picked case studies of good or bad practice, and discussion of various philosophers and sociologists.

@@

To objectively quantify the diversity of modern archaeological practice across a scale of relative hardness or softness, as an evaluation of its status as a science, and the place of this journal in context of other archaeology journals, I take a bibliometric approach. This approach is based on @fanelliBibliometricEvidenceHierarchy2013, who examined the hardness and softness of 12 disciplines using scholarly publication parameters. @fanelliBibliometricEvidenceHierarchy2013 found a spectrum of statistically significant variation in bibliometric variables from the physical to the social sciences, with papers tending at the softer end of the spectrum tending to have fewer co-authors, use less substantive titles, have longer texts, cite older literature, and have a higher diversity of sources. In @fanelliBibliometricEvidenceHierarchy2013's analysis harder sciences include Space Science, Physics, Chemistry, softer sciences include social sciences (Psychiatry, Psychology, Economics, Business and General Social Sciences), and the Humanities define the soft end of the spectrum. Following @fanelliBibliometricEvidenceHierarchy2013, I quantify the number of authors, length of article, relative title length, age of references, and diversity of references for a large sample of peer-reviewed journal articles.

@@

With many disciplines making increasing use of computationally intensive analyses in recent years there has been renewed interest in reproducibility [@levequeReproducibleResearchScientific2012]. In part, this is because computationally intensive research is difficult to communicate within the constraints of the methods section of a traditional journal article — the reader also needs the computer code written by the original authors, not just the article text. There is also the broader context of rising pressure to publish in prestigious journals and intense competition for funds that create strong incentives for malpractice in research [@edwards2017academic]. These two factors have led to widespread concerns of a reproducibility crisis in many fields [@Baker2016]. Estimates of scientific reproducibility in several fields confirm the extent of this problem. Empirical replications of 100 studies published in three psychology journals found that 36% of replications had statistically significant results, compared to 96% of the original studies [@OpenScienceCollaboration2015]. Similar empirical replications of large numbers of social science studies and experimental economics studies successfully replicated 61% and 62% of their target studies respectively [@camererEvaluatingReplicabilitySocial2018; @camererEvaluatingReplicabilityLaboratory2016].

@@

How does archaeology compare to these other fields in terms of reproducibility? Empirical reproducibility has long been valued in field archaeology. Throughout the history of archaeology, well-known sites have been repeatedly revisited to test old hypotheses with new evidence or methods, for example, Olduvai Gorge (Tanzania), Cahokia (USA), Çatalhöyük (Turkey), and Madjedbebe (Australia). Similarly among many experimental archaeologists, empirical reproducibility is a key concern, for example in lithic use-wear identification [@hayesLearningBlindTests2017] and the measurement of lithics [@pargeterReplicabilityLithicAnalysis2023]. The increasing availability of large digital datasets is pushing archaeology into unexplored areas [@bevan2015data], inviting questions about what reproducibility means for data intensive archaeological research. For example, to what extent does concern for reproducibility extend to computational reproducibility among archaeologists?

@@

library(patchwork)
(aer_plot_software + aer_plot_methods + aer_plot_repo) / (aer_plot_issues + aer_plot_issues_lang) + plot_annotation(tag_levels = 'A')
```

@@

Despite the relatively small number of reproducibility reviews reported on here, there are patterns of common issues that point to a small set of simple tasks authors can do that have high potential to increase reproducibility. The problem of incomplete materials can be tackled in several basic ways. First, authors should use a simple and logical folder structure to organise their code and data to be as self-contained as possible (ie. not requiring files outside of the project folder or any modifications by a reader to run the code). There are many excellent, simple, and widely-used templates that authors can choose from, e.g. @doi:10.1080/00031305.2017.1375986, @figueiredoSimpleKitUse2022, @Cookiecutter, @cooperGuideReproducibleCode2017 and @wilsonGoodEnoughPractices2017

@@

## What about qualitative archaeological research?

Although the introduction of reproducibility reviews signifies a growth in computational archaeology and a desire to evaluate the research products beyond the journal article, a very substantial amount of archaeological research is qualitative, with few or no numerical data involved in making knowledge claims about the human past. For example, many archaeological questions can be answered by the simple presence or absence of artefacts or features, or qualitative comparisons of basic artefact characteristics such as shape, colour, surface treatments, and raw material. Chaîne opératoire analyses by ceramic and lithic specialists is an especially productive area of archaeological research that often relies on comparison of narratives of manufacturing processes. Do these studies count as science and should they be held to standards of reproducibility in the same way that computationally intensive research is?

While there is a strong case for qualitative research counting as science, because it is often a systematic, empirical, and rigorous process of inquiry, the question of

@@

Computational reproducibility is not a panacea; it should not be used as a universally accepted criterion for research quality [@Leonelli2018]. Results that are fully reproducible can contain errors and fraud. It is no guarantee of code quality, or that statistics have been used appropriately (cf. @Vaiglova_2025; Crema 2025), or that data management is consistent with FAIR and CARE principles [@carrollOperationalizingCAREFAIR2021]. It is also time-consuming for authors to ensure their computational work can be reproduced, and for reviewers to evaluate. In a professional environment where job security and career progression is often associated with pressure to publish many high-impact papers, demands for authors to spend time on reproducibility, resulting in less time for publishing more papers, may seem frustrating. This may seem especially unfair to early career researchers on short-term contracts, who may feel the goalposts for career success are being moved and that they are being asked to do more work their graduate training has not prepared them for. This highlights the need for a culture shift among senior archaeological scientists to value reproducibility in hiring and promotion decision-making. Professors must contribute to this shift by nurturing a culture of reanalysis and reproducibility in their teaching, for example by using replication assignments and by training students in the most current best practices and tools for reproducible research, such as R and Python [@marwickHowUseReplication2020].
